---
title: "Bourbon Shiny App – Function Documentation"
author: "Chris Reddish"
date: "`r Sys.Date()`"
output: html_document
---

## 🔍 Overview

This R Markdown file documents the main functions developed for my final Shiny app project: **Bourbon Explorer**. The app allows users to explore a dataset of bourbons using interactive filters for Price, Rating, and ABV. The user can also compare multiple bourbons and see a ranked list based on a calculated score.

The dataset used is `BourbonFINAL.csv`, and the app uses `ggplot2`, `dplyr`, `tidyr`, and `shiny`.

---

## 🧠 Function 1: `filtered_data()`

**Purpose:**  
Filters the bourbon dataset using user-selected inputs for price, rating, and ABV. Also creates a calculated `Score` used for sorting.

**Inputs:**
- `input$price_range`: minimum and maximum price
- `input$rating_range`: rating slider
- `input$abv_range`: alcohol by volume range

**Output:**  
A filtered and scored dataset sorted by descending `Score`.

```{r eval=FALSE}
filtered_data <- reactive({
  bourbon_data %>%
    filter(
      Price >= input$price_range[1],
      Price <= input$price_range[2],
      Rating >= input$rating_range[1],
      Rating <= input$rating_range[2],
      Abv >= input$abv_range[1],
      Abv <= input$abv_range[2]
    ) %>%
    mutate(
      Proof = Abv * 2,
      Score = (Rating * 0.6) + (Proof * 0.2) - (Price * 0.2)
    ) %>%
    arrange(desc(Score))
})
```

---

## 📊 Function 2: `output$topPlot`

**Purpose:**  
Displays a horizontal bar chart showing the top N bourbons ranked by the custom `Score` metric calculated in `filtered_data()`.

**Inputs:**
- `input$top_n`: Number of bourbons to display in the plot  
- `filtered_data()`: The reactive dataset filtered and sorted by Score

**Output:**  
A `ggplot2` bar chart rendered using `renderPlot()`.

```{r eval=FALSE}
output$topPlot <- renderPlot({
  filtered_data() %>%
    slice_head(n = input$top_n) %>%
    ggplot(aes(x = reorder(Name, Score), y = Score)) +
    geom_col(fill = "darkred") +
    coord_flip() +
    labs(
      title = paste("Top", input$top_n, "Bourbons by Score"),
      x = "Bourbon", y = "Score"
    ) +
    theme_minimal()
})
```

---

## 📈 Function 3: `output$comparePlot`

**Purpose:**  
Creates a grouped bar chart comparing the selected bourbons across three attributes: Price, ABV, and Rating.

**Inputs:**
- `input$compare`: A character vector of selected bourbon names

**Output:**  
A grouped bar chart showing attribute values side-by-side for each bourbon.

```{r eval=FALSE}
output$comparePlot <- renderPlot({
  selected <- input$compare
  compare_df <- bourbon_data %>%
    filter(Name %in% selected) %>%
    select(Name, Price, Abv, Rating) %>%
    pivot_longer(cols = -Name)

  ggplot(compare_df, aes(x = name, y = value, fill = Name)) +
    geom_col(position = "dodge") +
    labs(
      title = "Comparison of Selected Bourbons",
      x = "Attribute", y = "Value"
    ) +
    theme_minimal()
})
```

---

## 📋 Function 4: `output$filteredTable`

**Purpose:**  
Displays the filtered data table so users can see exactly which bourbons match their criteria.

**Inputs:**
- No direct input (uses `filtered_data()`)

**Output:**  
A table rendered using `renderTable()`.

```{r eval=FALSE}
output$filteredTable <- renderTable({
  filtered_data()
})
```

---

## 💡 Insights and Challenges

### What I Learned
- How to use `reactive()` functions in Shiny
- How to link user inputs to dynamic filtering and plot rendering
- How to use `mutate()` to calculate a composite score
- How to handle user interface components like sliders and dropdowns

### Challenges Faced
- Designing a good filter experience while still supporting premium bourbons ($250+)
- Getting slider ranges to adapt dynamically to the dataset
- Debugging logic when no data met filter conditions

---

## 📝 Blog Summary

**What R Markdown Helps Me Accomplish:**  
This document helped me communicate exactly how my Bourbon Explorer app works. It combines written explanations with code examples in one place, making it easy to share, review, and improve.

**What I Created in This File:**  
A breakdown of each function used in the app, including what it does, what it takes in, what it returns, and how it's used in the UI.

---

## 🔗 GitHub Link

https://github.com/Cred1747/RStudio/blob/main/BourbonMarkdown.Rmd

