---
title: "Understanding Matrices and their Inverses in R"
author: "Chris Reddish"
date: "2025-02-10"
output: html_document
---
## Introduction
In this blog post, I will compute the determinant and inverse of matrices using R. The matrices given for this assignment are: Matrix A (10×10 square matrix) Matrix B (10×100 rectangular matrix):
```{r}
A <- matrix(1:100, nrow=10)  
B <- matrix(1:1000, nrow=10)
```
# Step 1: Checking Matrix Dimensions
For a matrix to be invertible, it must be square, meaning the number of rows must equal the number of columns. Let’s check the dimensions:

```{r}
print(paste("Dimensions of A:", paste(dim(A), collapse="x")))
print(paste("Dimensions of B:", paste(dim(B), collapse="x")))
```
Since B is not square, it cannot be inverted.
#Step 2: Compute the Determinant of A
The determinant helps determine if a matrix is invertible. A determinant of zero means the matrix is singular (not invertible).

```{r}
det_A <- det(A)
print(paste("Determinant of A:", det_A))
```
Step 3: Attempt to Compute Inverse
Now, I will attempt to compute the inverse using solve(A). If det(A) == 0, this will fail.
```{r}
if (det_A != 0) {
  inv_A <- solve(A)
  print("Inverse of A:")
  print(inv_A)
} else {
  print("Matrix A is singular and cannot be inverted.")
}
```
# Step 4: Verify Why B Cannot Be Inverted
```{r}
C <- matrix(sample(1:100, 100), nrow=10)

# Compute inverse of C
inv_C <- solve(C)
print("Inverse of C:")
print(inv_C)
```
Matrix C is square and has a nonzero determinant, making it invertible.