---
title: "Module #2 Introduction to basic R functions and Data Structures - assignment "
author: "Christopher Reddish"
date: "2025-01-20"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
 Your assignment, evaluate the following function call myMean. The data for this function called assignment. 

> assignment2 <- c(16, 18, 14, 22, 27, 17, 19, 17, 17, 22, 20, 22)
>myMean <- function(assignment2) { return(sum(assignment2)/length(assignment2)) }
> YOUR turn...

```{r}
assignment2 <- c(16, 18, 14, 22, 27, 17, 19, 17, 17, 22, 20, 22)
myMean <- function(assignment2) { return(sum(assignment2)/length(assignment2)) }
myMean(assignment2)
```
The myMean function calculates the mean of the input vector by summing all elements of the vector (sum(assignment2)) and dividing it by the length of the vector (length(assignment2)).
The function produces the correct mean of 19.25

To add more flexibilty we can make the the function data as x. This will now work with any assignment. 


```{r}
assignment2 <- c(16, 18, 14, 22, 27, 17, 19, 17, 17, 22, 20, 22)
assignment3 <- c(15, 18, 24, 21, 24, 13, 18, 21, 11, 21, 26, 29)
assignment4 <- c(18, 33, 25, 23, 11, 24, 29, 30, 11, 11, 10, 17,14)

#this works with any assignment
MyMean <- function(x) {return(sum(x) / length(x))}

MyMean(assignment2)
MyMean(assignment3)
MyMean(assignment4)

```


