---
title: "Module 9: Comparing Visualization Methods in R"
author: "Chris Reddish"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## **Introduction**
This report compares **three visualization techniques** in R: **Base R, Lattice, and ggplot2**. We analyze the `USArrests` dataset, which contains crime statistics for the 50 U.S. states. The dataset includes four key variables:
- **Murder:** The number of murders per 100,000 residents
- **Assault:** The number of assaults per 100,000 residents
- **Rape:** The number of rapes per 100,000 residents
- **UrbanPop:** The percentage of the population living in urban areas

We will explore how these variables relate and visualize key trends using different R visualization approaches.

```{r load-packages}
# Load necessary libraries
library(lattice)
library(ggplot2)
library(reshape2)
library(ggcorrplot) # For correlation heatmap
```

```{r load-data}
# Load dataset
USArrests <- read.csv("USArrests.csv", row.names = 1)
USArrests$State <- rownames(USArrests)
USArrests_long <- melt(USArrests, id.vars = "State", variable.name = "CrimeType", value.name = "Rate")
```

# **1️⃣ Comparison Across Visualization Methods**

## **Histogram of Murder Rates**
Understanding the distribution of **murder rates across states** helps identify which states have **higher crime levels** and if the data follows a normal distribution or is skewed.

### **Base R**
```{r base-hist}
hist(USArrests$Murder, 
     breaks = 10, col = "lightblue", border = "black",
     main = "Histogram of Murder Rates (Base R)",
     xlab = "Murder Rate (per 100,000 people)",
     ylab = "Number of States")

abline(v = mean(USArrests$Murder), col = "red", lwd = 2, lty = 2)
```

### **Lattice**
```{r lattice-hist}
histogram(~Murder, data = USArrests,
          main = "Histogram of Murder Rates (Lattice)",
          xlab = "Murder Rate",
          col = "blue")
```

### **ggplot2**
```{r ggplot-hist}
ggplot(USArrests, aes(x = Murder)) +
  geom_histogram(binwidth = 1, fill = "lightblue", color = "black") +
  geom_vline(aes(xintercept = mean(Murder)), color = "red", linetype = "dashed") +
  labs(title = "Histogram of Murder Rates (ggplot2)",
       x = "Murder Rate (per 100,000 people)",
       y = "Number of States") +
  theme_minimal()
```

**What we learn from these histograms:**
- The **majority of states** have lower murder rates, with a few having very high rates (right-skewed distribution).
- The red line (mean) shows that most states fall **below the national average**.
- ggplot2 provides a **more refined visualization**, while base R and lattice are effective but less visually polished.

## **Ordered Bar Chart: Murder Rate by State**
To compare crime levels across states, we rank them from lowest to highest murder rate.

```{r ggplot-bar}
ggplot(USArrests, aes(x = reorder(State, Murder), y = Murder, fill = Murder)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  scale_fill_gradient(low = "lightblue", high = "red") +
  labs(title = "Which States Have the Highest Murder Rates?",
       x = "State",
       y = "Murder Rate per 100,000 people") +
  theme_minimal()
```

**Key insights from this bar chart:**
- States with the highest murder rates are clearly highlighted.
- The **color gradient** makes it easier to see crime intensity differences.
- **Flipping the bars** ensures better readability.

## **Scatter Plot: Murder vs. Assault**
This visualization examines whether **higher assault rates correlate with higher murder rates**. 

### **Base R**
```{r base-scatter}
plot(USArrests$Assault, USArrests$Murder, 
     main = "Murder vs Assault (Base R)", 
     xlab = "Assault Rate",
     ylab = "Murder Rate",
     col = "blue", pch = 16)

abline(lm(Murder ~ Assault, data = USArrests), col = "red", lwd = 2)
```

### **Lattice**
```{r lattice-scatter}
xyplot(Murder ~ Assault, data = USArrests,
       main = "Murder vs Assault (Lattice)",
       xlab = "Assault Rate",
       ylab = "Murder Rate",
       panel = function(x, y, ...) {
           panel.xyplot(x, y, ...)
           panel.abline(lm(y ~ x), col = "red", lwd = 2)
       })
```

### **ggplot2**
```{r ggplot-scatter}
ggplot(USArrests, aes(x = Assault, y = Murder, color = UrbanPop)) +
  geom_point(size = 3, alpha = 0.7) +
  geom_smooth(method = "lm", color = "red") +
  scale_color_gradient(low = "blue", high = "red") +
  labs(title = "Does a Higher Assault Rate Mean More Murders?",
       x = "Assault Rate per 100,000",
       y = "Murder Rate per 100,000",
       color = "Urban Population") +
  theme_minimal()
```

**Key insights from scatter plots:**
- There is a **strong positive correlation** between **assault and murder rates**.
- More urbanized states (**higher UrbanPop values**) tend to have **higher crime rates**.
- The **ggplot2 version is visually superior**, showing clear trends and including a color gradient for extra insight.

## **Correlation Heatmap of Crime Variables**
```{r crime-correlation}
# Compute correlation matrix ensuring all values are included
crime_corr <- cor(USArrests[, c("Murder", "Assault", "Rape", "UrbanPop")], use = "everything", method = "pearson")

# Print correlation matrix to check if values are being computed
print(crime_corr)

# Create a correlation heatmap with proper spacing
ggcorrplot(crime_corr, method = "circle", type = "full", lab = TRUE, 
           title = "Correlation Between Crime Variables",
           outline.col = "white", colors = c("blue", "white", "red"))

```

**Key Takeaways:**
- **Murder and Assault are highly correlated**, meaning **states with more assaults also have more murders**.
- **Urban population has weaker correlations**, suggesting **crime isn’t just an urban problem**.
- The **heatmap provides a great summary of relationships between variables**.

# **Conclusion**
These **9 visualizations** demonstrate the strengths of **Base R, Lattice, and ggplot2**:
- **Base R** is simple but lacks customization.
- **Lattice** is useful for multi-panel comparisons but less flexible.
- **ggplot2** is the best for **high-quality, detailed visualizations**.

Overall, **ggplot2 offers the clearest insights** into crime trends and relationships, making it the best choice for data-driven decision-making.


